Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study

### 2020-05-28 manu_draft.Rmd
started bib file, outline slight update

---
title: "Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study"
output: 
  bookdown::word_document2:
    reference_docx: my_styles.docx
# bibliography: refs.bib
author: "Marcus W. Beck (mbeck@tbep.org), Perry de Valpine (pdevalpine@berkeley.edu), Rebecca Murphy (rmurphy@chesapeakebay.net), Ian Wren (ianw@sfei.org), Ariella Chelsky (ariellac@sfei.org), Melissa Foley (melissaf@sfei.org), David B. Senn (davids@sfei.org)"
urlcolor: blue
csl: ecology.csl
link-citations: true
---

@@

# extract bib entries from online
# bib_scrp('manu_draft.Rmd', 'refs.bib')
```

@@

The San Francisco Estuary (SFE) is the largest estuary on the Pacific Coast of North America and drains an area of approximately 200 thousand km$^2$ in the US state of California.  The estuary is subdivided into six sub-embayments having distinct geomorphological characteristics: Suisan Bay, San Pablo Bay, North Central Bay, Central Bay, South Bay,and Lower South Bay.  Major freshwater inputs enter the system through the Sacramento-San Joaquin Delta complex upstream of Suisun Bay, where the combined inflow from both rivers is approximately 28 km$^3$ per year.  The northern subembayments are river-dominated (salinity ranging from 0 to 15 ppt) with seasonal freshwater pulses primarily occurring in the winter and spring during the rainy season and subsequent snowmelt in the upper watershed.  The southern subembayments are marine-dominated with salinity ranging from 5 to 35 ppt depending on the tidal cycle, effluent discharge from wastewater treatment plants, and stormwater runoff.  South Bay is heavily urbanized and includes thirty-seven wastewater treatments plants that serve 7.2 million people.  Secondary treatment occurs at a majority of the treatment plants and the remaining effluent is discharged into the SFE.  Agricultural runoff from the upper watershed also contributes to nutrient loading in the SFE with the annual nutrient export estimated as approximately 30 thousand kg dy$^{-1}$ of nitrogen from the Delta.

@@

The SFE Research Program of the US Geological Survey has been collecting water quality data at fixed locations in the bay since 1969 (Cloern and Schraga, 2016; Schraga et al.2017).  This dataset is one of the longest continuous monitoring datasets available in a North American estuary and includes at least monthly sampling at 37 fixed sampling locations along a 145 km longitudinal transect from the lower Sacramento River in the Delta to lower South Bay.  Water quality constituent measurements available for each station include salinity, temperature, light attenuation, chlorophyll-a, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients. The current analysis focused on near-surface chlorophyll-a data collected biweekly to monthly along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36). Discrete chlorophyll concentrations at each station were determined by fluoremetric analysis with 90% acetone pigment extraction on GFF filters. Data collected between 1990-2017 were selected for analysis because it represented a suitable balance among three factors relevant to testing the statistical approaches, including sufficient length of record, consistent biweekly-monthly sampling, and a diverse set of stations covering the salinity gradient across multiple subembayments.  While sampling frequency varied somewhat over time or by station, all data were treated as unique time series within the statistical models (i.e., no spatial or temporal binning or averaging was done).

@@

Generalized Additive Models are an extension of linear models that can incorporate additive smoothed terms to model non-linear effects. The basic structural form is (Hastie and Tibshirani 1986):

@@

where the expected value of a dependent variable $Y$ conditional on predictors X$_1$ through X$_p$ is the sum of $s_j\left(.\right)$ smoothing functions for each predictor plus an interecept term $s_0$.  The smoothing functons are standardized and have expected values equal to zero so that $Es_j\left(X_j\right) = 0$.  Because GAMs are an extension of Generalized Linear Models, they can also include linear predictors in addition to the smoothed functions.  All methods herein use functions provided by the *mgcv* R package to fit GAMs using multiple parameter smoothing estimation methods (Wood, 2017).

The flexibility of the smoothing functions is a powerful component of GAMs that uses localized, non-linear spline functions to approximate the relationship of $Y$ with predictor $X_j$.  A spline is a piecewise function, such as a polynomial, that is connected at knots (or *k*-values) that smoothly join the separate pieces of the function.  Increasing the number of knots creates a smoother fit of the response variable against the predictor at the risk of overfitting, whereas choosing a small number of knots creates a more rigid function that may underfit the data.  There are several smoothing functions available in the *mgcv* package and the default method uses a thin-plate spline as a computationally efficient approach to estimating the smoother (Wood, 2003).  The thin-plate spline can provide a balance between over- and under-fitting to the bivariate smoother by identifying the optimal number of knots through generalized cross validation (GCV).  In addition to providing an overall summary statistic of model performance, the GCV optimization informs the number of knots for the spline by balancing the fit subject to a smoothing penalty. This frees the user from manually selecting knots, and therefore, the degree of smoothness modelled by the function. However, the *mgcv* package requires the user to specify an upper limit as a reasonable expection for the number of knots applied to each spline.

@@

To evaluate GAMs with SFE chloropyll data, we followed an approach that built on past work by the Chesapeake Bay Program (Murphy et al. 2019), including incorporation and adaptation of their hierarchical model structure to evaluate long-term time series.  These hierarchical models vary the functional form for time in relation to chlorophyll by adding smoothers and linear predictors with increasing flexibility in the GAM functions.  This is a powerful approach for time series analysis because separate components can be explicitly modelled with different smoothing terms and the results can be idendependently evaluated and further tested from the model output.  Varying degrees of model complexity also provides a balance between descriptive power and computational efficiency, where simpler models may be preferred in routine reporting or more complex models may be preferred for hypothesis testing.  We present four GAM structures, following a similar format as Murphy et al. (2019).  The notation for gam3, gam4, and gam5 are not used because these describe models in Murphy et al. (2019) that were developed for purposes specific to Chesapeake Bay.

@@

The link functions $g()$ for each model were chosen as either logarithmic transformations (base 10) or best estimates of Box-Cox power transformations for chlorophyll.  A comparison of GAM perforamcne between transformation methods has not yet been explored in the application of GAMs to long-term water quality monitoring data.  The logarithmic transformation is commonly used for chlorophyll to approximate a log-normal distribution of the response variable. However, alternative transformations could provide improved model fits and reduced uncertainty estimates in the fitted parameters by better satisying assumptions for GAMs.  The Box-Cox method was also used to transform chlorophyll as a comparison to the logarithmic transformation (Box and Cox 1964).  The Box-Cox method requires an estimate of the parameter $\lambda$, whereby the the optimal value is based on a minimization function of a log-likelihood profile vector.  The profile vector for each chlorophyll time series was obtained using the *boxcox* funtion from the MASS package (Venables and Ripley 2002) to evaluate log-likelihood estimates of $\lambda$ across the range -4 to 4.  Once the optimal $\lambda$ value was identified, chlorophyll was transformed using the following power transformation (Box and Cox 1964):

@@

where $g()$ is the transformation function for the appropriate GAM at the esitmated $\lambda$ value. Note that in some cases the Box-Cox transformation is a logarithmic transformation if $\lambda$ is estimated as zero.  The *BoxCox* function from the forecast package (Hyndman et al. 2020) in R was used to transform the chlrophyll time series with the optimal $\lambda$ value.  Equation \@ref(eq:boxcox) is only appropriate for positive values of a response variable, as for chlorophyll.

@@

Each of the GAM structures described in equations \@ref(eq:gamzero) to \@ref(eq:gamsix) provide different approaches to quantify long-term variation in annual and seasonal changes.  Although useful as a general approach to view functional forms for different time series, additional information regarding whether an observed change is statistically significant can be useful to inform environmental management decisions.  We present two methods for performing trend tests on model results that address different changes that may be of management interest.  First, a method for evaluating the percent change in annual averages between two time periods is briefly summarized.  This approach is similar to that presented in Murphy et al. (2019).  Second, a method for evaluating means and confidence intervals from GAM results for seasonal periods is presented, including an estimate of direction and rate of seasonal change over time.

The ability to compare time periods of interest has importance for evaluating whether a significant change has occurred in relation to perceived trends in observed data.  For example, the noticeable change in SFE water quality in South Bay starting from the 1990s and persisting through 2005 was of considerable interest by environmental managers given the apparent departure from long-term conditions to a state that suggested a trend towards more eutrophic conditions.  However, statistical methods to evaluate this change that were able to quantify the temporal variation and to provide a statistically robust estimate of the change were not available.  The use of GAMs in parallel with hypothesis testing can be used for this purpose. Methods proposed in Murphy et al. (2019) and similarly applied in Yang and Moyer (2020) were used and are briefly summarized herein.

For two time periods of interest, a trend can be estimated as the difference in means from the GAM results between the periods, the percent change, and an appropriate p-value descibing the probability that the change is due to change alone.  Murphy et al. (2019) defined the two time periods as "base" and "test" periods, where the former represents a baseline condition earlier in the time series as a basis of comparison to the more recent "test" period.  It was further suggested the the base and test periods include a minimum of two years if evaluating years at the beginning and end of a time series to account for instability of GAM estimates at the edges of the independent variable space.  The differences are estimated as:

@@

where the estimated value of chlorophyll at year $i$ for the predetermined season is a function of the predicted seasonal mean $\beta_0 +\beta_1 year_i$ plus deviation of the seasonal mean from the trend ($b_i$) and devation of the seasonal mean from the true mean ($S_i$).  Both the error terms are assumed to be normally distributed with mean zero and variance for $b_i$ equal to an unknown term and vraiance for $\epsilon_i$ equal to the squared standard error from the seasonal chloropyll averages.  The R package *mixmeta* was used to estimate trends in the seasonal averages over time (Sera et al. 2019).  Results appear similar to those from regression analysis (e.g., slope and intercept estimates), although parameter uncertainty is more accurately estimated with mixed meta-analysis.

>>>>

---
title: "Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study"
output: 
  bookdown::word_document2:
    reference_docx: my_styles.docx
bibliography: refs.bib
author: "Marcus W. Beck (mbeck@tbep.org), Perry de Valpine (pdevalpine@berkeley.edu), Rebecca Murphy (rmurphy@chesapeakebay.net), Ian Wren (ianw@sfei.org), Ariella Chelsky (ariellac@sfei.org), Melissa Foley (melissaf@sfei.org), David B. Senn (davids@sfei.org)"
urlcolor: blue
csl: ecology.csl
link-citations: true
---

@@

# # extract bib entries from online
# bib_scrp('manu_draft.Rmd', 'refs.bib')
```

@@

The San Francisco Estuary (SFE) is the largest estuary on the Pacific Coast of North America and drains an area of approximately 200 thousand km$^2$ in the US state of California.  The estuary is subdivided into six sub-embayments having distinct geomorphological characteristics: Suisun Bay, San Pablo Bay, North Central Bay, Central Bay, South Bay,and Lower South Bay.  Major freshwater inputs enter the system through the Sacramento-San Joaquin Delta complex upstream of Suisun Bay, where the combined inflow from both rivers is approximately 28 km$^3$ per year.  The northern subembayments are river-dominated (salinity ranging from 0 to 15 ppt) with seasonal freshwater pulses primarily occurring in the winter and spring during the rainy season and subsequent snowmelt in the upper watershed.  The southern subembayments are marine-dominated with salinity ranging from 5 to 35 ppt depending on the tidal cycle, effluent discharge from wastewater treatment plants, and stormwater runoff.  South Bay is heavily urbanized and includes thirty-seven wastewater treatments plants that serve 7.2 million people.  Secondary treatment occurs at a majority of the treatment plants and the remaining effluent is discharged into the SFE.  Agricultural runoff from the upper watershed also contributes to nutrient loading in the SFE with the annual nutrient export estimated as approximately 30 thousand kg dy$^{-1}$ of nitrogen from the Delta.

@@

The SFE Research Program of the US Geological Survey has been collecting water quality data at fixed locations in the bay since 1969 [@Cloern16;@Schraga20].  This dataset is one of the longest continuous monitoring datasets available in a North American estuary and includes at least monthly sampling at 37 fixed sampling locations along a 145 km longitudinal transect from the lower Sacramento River in the Delta to lower South Bay.  Water quality constituent measurements available for each station include salinity, temperature, light attenuation, chlorophyll-a, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients. The current analysis focused on near-surface chlorophyll-a data collected biweekly to monthly along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36). Discrete chlorophyll concentrations at each station were determined by fluoremetric analysis with 90% acetone pigment extraction on GFF filters. Data collected between 1990-2017 were selected for analysis because it represented a suitable balance among three factors relevant to testing the statistical approaches, including sufficient length of record, consistent biweekly-monthly sampling, and a diverse set of stations covering the salinity gradient across multiple subembayments.  While sampling frequency varied somewhat over time or by station, all data were treated as unique time series within the statistical models (i.e., no spatial or temporal binning or averaging was done).

@@

Generalized Additive Models are an extension of linear models that can incorporate additive smoothed terms to model non-linear effects. The basic structural form is [@Hastie90]:

@@

where the expected value of a dependent variable $Y$ conditional on predictors X$_1$ through X$_p$ is the sum of $s_j\left(.\right)$ smoothing functions for each predictor plus an interecept term $s_0$.  The smoothing functons are standardized and have expected values equal to zero so that $Es_j\left(X_j\right) = 0$.  Because GAMs are an extension of Generalized Linear Models, they can also include linear predictors in addition to the smoothed functions.  All methods herein use functions provided by the *mgcv* R package to fit GAMs using multiple parameter smoothing estimation methods [@Wood17].

The flexibility of the smoothing functions is a powerful component of GAMs that uses localized, non-linear spline functions to approximate the relationship of $Y$ with predictor $X_j$.  A spline is a piecewise function, such as a polynomial, that is connected at knots (or *k*-values) that smoothly join the separate pieces of the function.  Increasing the number of knots creates a smoother fit of the response variable against the predictor at the risk of overfitting, whereas choosing a small number of knots creates a more rigid function that may underfit the data.  There are several smoothing functions available in the *mgcv* package and the default method uses a thin-plate spline as a computationally efficient approach to estimating the smoother [@Wood03].  The thin-plate spline can provide a balance between over- and under-fitting to the bivariate smoother by identifying the optimal number of knots through generalized cross validation (GCV).  In addition to providing an overall summary statistic of model performance, the GCV optimization informs the number of knots for the spline by balancing the fit subject to a smoothing penalty. This frees the user from manually selecting knots, and therefore, the degree of smoothness modelled by the function. However, the *mgcv* package requires the user to specify an upper limit as a reasonable expection for the number of knots applied to each spline.

@@

To evaluate GAMs with SFE chloropyll data, we followed an approach that built on past work by the Chesapeake Bay Program [@Murphy19], including incorporation and adaptation of their hierarchical model structure to evaluate long-term time series.  These hierarchical models vary the functional form for time in relation to chlorophyll by adding smoothers and linear predictors with increasing flexibility in the GAM functions.  This is a powerful approach for time series analysis because separate components can be explicitly modelled with different smoothing terms and the results can be idendependently evaluated and further tested from the model output.  Varying degrees of model complexity also provides a balance between descriptive power and computational efficiency, where simpler models may be preferred in routine reporting or more complex models may be preferred for hypothesis testing.  We present four GAM structures, following a similar format as @Murphy19.  The notation for gam3, gam4, and gam5 are not used because these describe models in @Murphy19 that were developed for purposes specific to Chesapeake Bay.

@@

The link functions $g()$ for each model were chosen as either logarithmic transformations (base 10) or best estimates of Box-Cox power transformations for chlorophyll.  A comparison of GAM perforamcne between transformation methods has not yet been explored in the application of GAMs to long-term water quality monitoring data.  The logarithmic transformation is commonly used for chlorophyll to approximate a log-normal distribution of the response variable. However, alternative transformations could provide improved model fits and reduced uncertainty estimates in the fitted parameters by better satisying assumptions for GAMs.  The Box-Cox method was also used to transform chlorophyll as a comparison to the logarithmic transformation (Box and Cox 1964).  The Box-Cox method requires an estimate of the parameter $\lambda$, whereby the the optimal value is based on a minimization function of a log-likelihood profile vector.  The profile vector for each chlorophyll time series was obtained using the *boxcox* funtion from the MASS package [@Venables02 to evaluate log-likelihood estimates of $\lambda$ across the range -4 to 4.  Once the optimal $\lambda$ value was identified, chlorophyll was transformed using the following power transformation [@Box64]:

@@

where $g()$ is the transformation function for the appropriate GAM at the esitmated $\lambda$ value. Note that in some cases the Box-Cox transformation is a logarithmic transformation if $\lambda$ is estimated as zero.  The *BoxCox* function from the forecast package [@Hyndman20] in R was used to transform the chlrophyll time series with the optimal $\lambda$ value.  Equation \@ref(eq:boxcox) is only appropriate for positive values of a response variable, as for chlorophyll.

@@

Each of the GAM structures described in equations \@ref(eq:gamzero) to \@ref(eq:gamsix) provide different approaches to quantify long-term variation in annual and seasonal changes.  Although useful as a general approach to view functional forms for different time series, additional information regarding whether an observed change is statistically significant can be useful to inform environmental management decisions.  We present two methods for performing trend tests on model results that address different changes that may be of management interest.  First, a method for evaluating the percent change in annual averages between two time periods is briefly summarized.  This approach is similar to that presented in @Murphy19.  Second, a method for evaluating means and confidence intervals from GAM results for seasonal periods is presented, including an estimate of direction and rate of seasonal change over time.

The ability to compare time periods of interest has importance for evaluating whether a significant change has occurred in relation to perceived trends in observed data.  For example, the noticeable change in SFE water quality in South Bay starting from the 1990s and persisting through 2005 was of considerable interest by environmental managers given the apparent departure from long-term conditions to a state that suggested a trend towards more eutrophic conditions.  However, statistical methods to evaluate this change that were able to quantify the temporal variation and to provide a statistically robust estimate of the change were not available.  The use of GAMs in parallel with hypothesis testing can be used for this purpose. Methods proposed in @Murphy19 and similarly applied in @Yang20 were used and are briefly summarized herein.

For two time periods of interest, a trend can be estimated as the difference in means from the GAM results between the periods, the percent change, and an appropriate p-value descibing the probability that the change is due to change alone.  @Murphy19 defined the two time periods as "base" and "test" periods, where the former represents a baseline condition earlier in the time series as a basis of comparison to the more recent "test" period.  It was further suggested the the base and test periods include a minimum of two years if evaluating years at the beginning and end of a time series to account for instability of GAM estimates at the edges of the independent variable space.  The differences are estimated as:

@@

where the estimated value of chlorophyll at year $i$ for the predetermined season is a function of the predicted seasonal mean $\beta_0 +\beta_1 year_i$ plus deviation of the seasonal mean from the trend ($b_i$) and devation of the seasonal mean from the true mean ($S_i$).  Both the error terms are assumed to be normally distributed with mean zero and variance for $b_i$ equal to an unknown term and variance for $\epsilon_i$ equal to the squared standard error from the seasonal chloropyll averages.  The R package *mixmeta* was used to estimate trends in the seasonal averages over time [@Sera19].  Results appear similar to those from regression analysis (e.g., slope and intercept estimates), although parameter uncertainty is more accurately estimated with mixed meta-analysis.
